---
title: Conditional recoding of values in a data frame with recode_if()
author: Garrick Aden-Buie
date: '2018-07-17'
slug: recode-if
categories: []
tags:
  - R
  - Script
  - Tips
hero_bg: "/img/hero/victor-garcia-645096-unsplash.jpg"
---

```{r settings, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# knit_print.tbl_df = function(x, ...) {
#     res = paste(c("", "", knitr::kable(x)), collapse = "\n")
#     knitr::asis_output(res)
# }

library(dplyr)
example_data <- tibble::tribble(
   ~id,                     ~name,                                ~job, ~code,
  8350,           "Nolia Osinski", "Research officer, political party", "XDR",
  1045,         "Dr. Gorge Green",              "Sports administrator", "AWG",
  3839,         "Bernie O'Reilly",                       "Ship broker", "AWG", #CLP
  5643,    "Brionna Kuphal-Fadel",              "Sports administrator", "_MXN",
  1891,    "Patsy Lind-Halvorson",                "Purchasing manager", "SEK",
  1506,              "Samual Fay",                        "Ergonomist", "999", #SYP
  5757,       "Gabriela Kassulke",                "Engineer, drilling", "HTG",
  6702,            "Dr. Ty Lynch",           "Physiological scientist", "_MKD",
  2702, "Abigale Senger-Schimmel",                  "Engineer, energy", "BTN",
  3622,   "Regis Stracke-Bartell",                "Surveyor, minerals", "THB"
)
example_data_orig <- example_data
```

Suppose your data set has a column that contains errors in a small number of values.
Additionally, you know that the location of these errors can be determined based on another column, such as an ID.
Finally, you'd like to make it extremely clear in your code which values were changed and to what so that there is no confusion among your collaborators (or your future self).

For example, consider the following completely made up data containing a few issues in the `code` column.

```{r}
example_data
```

There are three issues in the `code` column that need to be rectified:

1. One value is `999` when it should be `SYP` (row 6).

1. The `AWG` value for Bernie O'Reilly (row 3) was miscoded at the time of data entry and should be `CLP`.

1. Two of the values for recently added people start with a `_` that should be removed (rows 4 and 8). Here, recently added means `id >= 5000`.

Note that in this context, because we are changing values from the source data, we want to be especially clear in our code that we are recoding `"AWG"` as `"CLP"` for person 3989.

To do this, we'll build on the dplyr function `recode()` to create a version called `recode_if()` that only applies the recoding when a secondary condition is met.

## Standard `recode()`

The first issue can be fixed with a standard use of `recode()` from dplyr.

```{r}
example_data <- example_data %>% 
  mutate(code = recode(code, "999" = "SYP"))
example_data
```

`recode()` is a vectorized version of `switch()`.
In the above example, `recode()` works by taking any value of `code` equal to `999` and _recoding it_ as `"SYP"`.

## `recode_if()`

For the second two issues, we need to condition the value updating on a second column.
By this, we mean that we can't use the value of `code` directly to choose which value to update -- for example, we need to change the value of `code` when it equals `AWG`, but only for `id == 3839`.

To do this we introduce a simple function called `recode_if()` that provides a wrapper around `if_else()` and `recode()`.

```{r}
recode_if <- function(x, condition, ...) {
  if_else(condition, recode(x, ...), x)
}
```

Then we apply this function to change the value of `code` to `"CLP"` for the person with `id == 3839`.

```{r}
example_data <- example_data %>%
  mutate(code = recode_if(code, id == 3839, "AWG" = "CLP"))

example_data
```

And finally, we strip the leading `"_"` character for people with `id >= 5000` using `recode_if()`.

```{r}
example_data <- example_data %>%
  mutate(code = recode_if(code, id >= 5000, "_MXN" = "MXN", "_MKD" = "MKD"))

example_data
```

## Comparison

The example here was contrived, but it's not uncommon to need to change the value in a single row/column pair in a data frame during cleaning.
`recode()` and `recode_if()` are two methods that are useful, but there are others.
In our opinion, the explict mapping of old values to new values in `recode()` and `recode_if()` makes the code clearer and easier to understand from a distance.

Another option is to use `if_else()` directly

```{r}
example_data_orig %>% 
  mutate(
    code = if_else(code == "999", "SYP", code),
    code = if_else(id == 3839, "CLP", code),
    code = if_else(id >= 5000,
                   if_else(code %in% c("_MXN", "_MKD"), 
                           sub("^_", "", code),
                           code),
                   code)
  )
```

but the nested `if_else()` statements gets unwieldy when there are multiple values.

A third option is to use `case_when()`, as in

```{r}
example_data_orig %>% 
  mutate(
    code = case_when(
      code == "999" ~ "SYP",
      id == 3839 ~ "CLP",
      id >= 5000 & code == "_MXN" ~ "MXN",
      id >= 5000 & code == "_MKD" ~ "MKD",
      TRUE ~ code
    )
  )
```

but the repeated `id >= 5000` filter makes the recoding less clear.
(Note that this a contrived example, so pretend this filter is neccessary and unavoidable.)

---

Cover photo by: Photo by [Victor Garcia](https://unsplash.com/photos/0yL6nXhn0pI?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText) on [Unsplash](https://unsplash.com/search/photos/grid?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText)