---
title: Encrypting and Sharing Data Containing PHI
author: 
  - Garrick Aden-Buie
  - Travis Gerke
date: '2019-02-22'
slug: encrypting-and-share-phi
categories:
  - R
tags:
  - R
description: "Using the encryptr package to securely share data with collaborators"
weight: 20
hero_bg: "/img/hero/corey-agopian-106075-unsplash.jpg"
hero_credit: "[Corey Agopian](https://unsplash.com/photos/-yhpJ0BwLcU?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText) on Unsplash"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  class.source = "code-source",
  class.output = "code-output"
)
# Run this post in a temp directory
knitr::opts_knit$set(root.dir = tempdir())

if (interactive()) setwd(tempdir())
```

<!-- Links -->

#### Links (will remove)

[encryptr]: https://github.com/SurgicalInformatics/encryptr
[sodium]:   https://cran.r-project.org/web/packages/sodium/vignettes/intro.html
[cyphr]:    https://github.com/ropensci/cyphr
[r-secret]: https://github.com/gaborcsardi/secret

- [encryptr]
- [sodium]
- [cyphr]
- [r-secret]

## Skeleton of Example

```{r library-encryptr}
# remotes::install_github("SurgicalInformatics/encryptr")
library(encryptr)
```

```{r library-general}
library(tidyverse)
```

Three people who need to share data:

* **Org**: The **Owner** or organization sharing the data

* **Aiyana**: A fully-cleared collaborator

* **Zane**: A collaborator who cannot access PHI

To set up, we create a public/private key pair for each person.

```{r create-keypairs}
# I have to do this manually because `encryptr` doesn't expose the
# password argument in openssl::write_pem()

people <- c("org", "aiyana", "zane")
for (person in people) {
  # straight out of ?openssl::rsa_keygen()
  key <- openssl::rsa_keygen()
  pubkey <- as.list(key)$pubkey
  openssl::write_pem(key, paste0(person, "_private_key"))
  openssl::write_pem(pubkey, paste0(person, "_public_key"))
}

dir()
```

### The owner has the data and encrypts it

The owner has some data about breast cancer patients.

```{r data-create, include=FALSE}
patient_data <- 
tibble::tribble(
  ~medical_record_number,                      ~primary_site,     ~date_dx,
                  984584,                "C503 (BREAST LIQ)", "2009-02-01",
                  534924,                "C509 (BREAST NOS)", "2012-04-13",
                  624219,                "C509 (BREAST NOS)", "2010-04-25",
                  112743,                "C503 (BREAST LIQ)", "2014-12-03",
                  261730,           "C383 (MEDIASTINUM NOS)", "2014-08-27",
                  194777, "C508 (BREAST OVERLAPPING LESION)", "2013-03-18",
                  446950,               "C541 (ENDOMETRIUM)", "2016-03-28",
                  751595,                "C504 (BREAST UOQ)", "2012-08-18",
                  385615,                "C504 (BREAST UOQ)", "2011-01-08",
                  761953,                "C504 (BREAST UOQ)", "2012-08-08"
  )
```

```{r data-view}
patient_data
```

They need to hide the Medical Record Number and Date of Diagnosis.

```{r data-encrypted}
patient_encrypted <- 
  patient_data %>% 
  encrypt(
    medical_record_number,
    date_dx,
    public_key_path = "org_public_key"
  )

patient_encrypted
```

Can Aiyana or Zane see the data?

```{r data-decrypt-try, error=TRUE}
patient_encrypted %>% 
  decrypt(
    medical_record_number,
    date_dx,
    private_key_path = "aiyana_private_key"
  )

patient_encrypted %>% 
  decrypt(
    medical_record_number,
    date_dx,
    private_key_path = "zane_private_key"
  )
```

Nope.
But the Org can!

```{r data-decrypt-org}
patient_encrypted %>% 
  decrypt(
    medical_record_number,
    date_dx,
    private_key_path = "org_private_key"
  )
```

### Aiyana wants the data

We encrypt for Aiyana to receive the data.

```{r data-encrypt-for-a}
patient_encrypt_A <- 
  patient_data %>% 
  encrypt(
    medical_record_number, date_dx,
    public_key_path = "aiyana_public_key"
  )

patient_encrypt_A
```

```{r data-decrypt-try-a, error=TRUE}
patient_encrypt_A %>% 
  decrypt(
    medical_record_number,
    date_dx,
    private_key_path = "aiyana_private_key"
  )

patient_encrypt_A %>% 
  decrypt(
    medical_record_number,
    date_dx,
    private_key_path = "zane_private_key"
  )
```