---
title: "About"
description: "Meet Team Gerke"
draft: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(purrr)
library(glue)
```

```{r}
links <- list(
  "Travis Gerke" = list(
    email = "Travis.Gerke@moffitt.org",
    github = "tgerke",
    twitter = "travisgerke",
    link = "travisgerke.com"
  ),
  "Jordan Creed" = list(
    email = "Jordan.Creed@moffitt.org",
    github = "jhcreed"
  ),
  "Garrick Aden-Buie" = list(
    email = "Garrick.Aden-Buie@moffitt.org",
    github = "gadenbuie",
    twitter = "grrrck",
    link = "garrickadenbuie.com"
  )
)

link_images <- list(
  email = "far fa-envelope-open",
  twitter = "fab fa-twitter",
  github = "fab fa-github",
  link = "fas fa-link"
)

link_prefix <- list(
  email = "mailto:",
  twitter = "https://twitter.com/",
  github = "https://github.com/",
  link = "https://"
)

make_table <- function(name) {
  this_links <- links[[name]]
  if (!length(this_links)) return("")
  rows <- purrr::imap(this_links, ~ glue::glue(
    '<tr>\n<td><i class="{link_images[[.y]]}"></i></td>\n',
    '<td><a href="{link_prefix[[.y]]}{.x}">{.x}</a></td>\n</tr>')
  )
  rows <- paste(rows, collapse = "\n")
  paste(
    '<table style="width: 90%">',
    rows,
    '</table>',
    sep = "\n"
  )
}

headshot_image <- function(img) {
  paste0('<div class="headshot-container"><img class="headshot w90" src="', img, '"></div>')
}
```

:::::: {.row .gutters .about-card}
:::: {.col .col-6}
## Travis Gerke

`r headshot_image("/img/headshot-gerke.jpg")`

`r make_table("Travis Gerke")`

::::
:::: {.col .col-6}

Our group's research focus lies in integrative molecular cancer epidemiology, and is driven by concurrent interests in biostatistical methods development, computational biology, and modern data science practice.
In addition to collaborative efforts in the domain of classical cancer epidemiology, our group is involved in several efforts to streamline processing, analysis, and epidemiologic translation of large-scale genomic databases such as The Cancer Genome Atlas (TCGA) and the Gene Expression Omnibus.
An emerging research interest in clinical epidemiology applies contemporary methods in causal inference to existing clinical trial data in order to discover optimal treatment sequences.

A cross-cutting theme across the lab is the use of R software for data science.
We actively develop and maintain several R packages and user interfaces (e.g., Shiny applications) that aim to alleviate computational bottlenecks associated with conducting molecular epidemiology research.

For a full list of publications and other academic activities, please see my [complete CV](http://travisgerke.com/cv/).

::::
::::::


:::::: {.row .gutters}
:::: {.col .col-6 .about-card}
## Jordan Creed

`r headshot_image("/img/headshot-jordan.png")`

`r make_table("Jordan Creed")`

Jordan is a Research Data Analyst...

::::

:::: {.col .col-6 .about-card}
## Garrick <span style="white-space: nowrap">Aden-Buie</span>

`r headshot_image("/img/headshot-garrick.jpg")`

`r make_table("Garrick Aden-Buie")`

Garrick is a Research Data Analyst and expert R developer and educator who is passionate about reproducible research and applying software carpentry best practices within scientific computing.
He is an ABD doctoral candidate in the department of Industrial Engineering at the University of South Florida, where he built technological and data driven solutions for smart, healthy aging.
While new to the world of bioinformatics, he's excited to be bringing his knowledge of data science dev ops and analytics to Gerke Lab and computational biology.
::::
::::::